# Cyclin A1 Expression in pT1 Urothelial Carcinoma of Bladder
## Opening & attaching dataset and loading functions & libraries
```{r open}
ca1 <- read.csv("../ca1.csv")
attach(ca1)
library(psych) # For descriptive statistics
library(survival) # For Survival Analyses
source("../standardAnalysis.R")
source("../SurvivalPlot.R")
```
## Describing the Data
```{r describe}
# pT1 at Biopsy
table.prop(pt.biopsy)
# CIS at Biopsy
table.prop(cis.biopsy) 
# Initial Treatment
table.prop(treatment)
# Recurrence ratio
table.prop(recurrence)
# Distribution of TNM stages
table.prop(dx.final) 
# Describing the number of recurrence episodes
descriptive(recurrence.number) 
# Recurrence episodes by groups
table.prop(recurrence.groups) 
# Describing the % of cyclin A1 expression
descriptive(ca1.extension)
# % Cyclin A1 Positivity
table.prop(ca1.positive.10)
```

# Table 1
```{r}
# Cyclin A1 by pT at Biopsy
compare.fisher(pt.biopsy, ca1.positive.10)
# Cyclin A1 by CIS at Biopsy
compare.fisher(cis.biopsy, ca1.positive.10)
# Cyclin A1 by Recurrence
compare.fisher(recurrence, ca1.positive.10)
# Cyclin A1 by Progression
compare.fisher(progression, ca1.positive.10) 
# Adjusting P value using Hommel's correction
table1.raw <- c(0.01864, 0.7189, 1, 0.00325)
table1.hommel <- p.adjust(table1.raw, method = c("hommel"))
table1.adjusted <- cbind(table1.raw, table1.hommel)
matrix(data = table1.adjusted, nrow = 4, ncol = 2,
       dimnames = list(c("pT at Biopsy", "CIS at Biopsy", "Recurrence", "Progression"),
                       c("Raw P values", "Adjusted P values")))
```

## Estimating Survival Functions for Tumor Recurrence
```{r SurvivalRecurrence}
# Defining survival object
surv.recurrence <- with(ca1, Surv(time.event, recurrence))
# pT1 and recurrence
pt1.recurrence <- survfit(surv.recurrence ~ pt.biopsy, data = ca1)
# CIS and recurrence
cis.recurrence <- survfit(surv.recurrence ~ cis.biopsy, data = ca1)
# Initial treatment and recurrence
treatment.recurrence <- survfit(surv.recurrence ~ treatment, data = ca1)
# Cyclin A1 and recurrence
ca1.recurrence <- survfit(surv.recurrence ~ ca1.positive.10, data = ca1)
```

### pT Biopsy and Tumor Recurrence (Figure 2A)
```{r Figure2A, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.recurrence ~ pt.biopsy)
# Survival curves
survival.plot(x = pt1.recurrence, title = "pT at Biopsy and Tumor Recurrence",
p = 0.02, levels = c("pT1a", "pT1b"))
```

### CIS and Tumor Recurrence (Figure 2B)
```{r Figure2B, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.recurrence ~ cis.biopsy)
# Survival curves
survival.plot(x = cis.recurrence, title = "CIS at Biopsy and Tumor Recurrence",
p = 0.32, levels = c("Absent", "Present"))
```

### Initial Treatment and Tumor Recurrence (Figure 2C)
```{r Figure2C, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.recurrence ~ treatment)
# Survival curves
survival.plot(x = treatment.recurrence, title = "Initial Treatment at Biopsy and Tumor Recurrence", p = 0.47, levels = c("TURB+BCG", "TURB+MC", "TURB alone"))
```

### Cyclin A1 and Tumor Recurrence (Figure 2D)
```{r Figure2D, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.recurrence ~ ca1.positive.10)
# Survival curves
survival.plot(x = ca1.recurrence, title = "Cyclin A1 and Tumor Recurrence", p = 0.68, position = "topright", levels = c("Negative", "Positive"))
```

### Tumor Recurrence (All Figures Combined)
```{r Figure2AD, fig.height = 11, fig.width = 13}
par(mfrow = c(2,2))
# Figure 2A
survival.plot(x = pt1.recurrence, title = "pT at Biopsy and Tumor Recurrence",
p = 0.02, levels = c("pT1a", "pT1b"))
# Figure 2B
survival.plot(x = cis.recurrence, title = "CIS at Biopsy and Tumor Recurrence",
p = 0.32, levels = c("Absent", "Present"))
# Figure 2C
survival.plot(x = treatment.recurrence, title = "Initial Treatment at Biopsy and Tumor Recurrence", p = 0.47, levels = c("TURB+BCG", "TURB+MC", "TURB alone"))
# Figure 2D
survival.plot(x = ca1.recurrence, title = "Cyclin A1 and Tumor Recurrence", p = 0.68, position = "topright", levels = c("Negative", "Positive"))
```

## Estimating Survival Functions for Tumor Progression
```{r SurvivalProgression}
# For tumor progression
surv.progression <- with(ca1, Surv(time.event, progression))
# pT1 and progression
pt1.progression <- survfit(surv.progression ~ pt.biopsy, data = ca1)
# CIS and progression
cis.progression <- survfit(surv.progression ~ cis.biopsy, data = ca1)
# Initial treatment and progression
treatment.progression <- survfit(surv.progression ~ treatment, data = ca1)
# Cyclin A1 and progression
ca1.progression <- survfit(surv.progression ~ ca1.positive.10, data = ca1)
```

### pT at Biopsy and Tumor Progression (Figure 2E)
```{r Figure2E, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.progression ~ pt.biopsy)
# Survival curves
survival.plot(x = pt1.progression, title = "pT at Biopsy and Tumor Progression", p = 1.11e-06, levels = c("pT1a", "pT1b"))
```

### CIS at Biopsy and Tumor Progression (Figure 2F)
```{r Figure2F, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.progression ~ cis.biopsy)
# Survival curves
survival.plot(x = cis.progression, title = "CIS at Biopsy and Tumor Progression", p = 0.03, levels = c("Absent", "Present"), position = "bottomright")
```

### Initial Treatment and Tumor Progression (Figure 2G)
```{r Figure2G, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.progression ~ treatment)
# Survival curves
survival.plot(x = treatment.progression, title = "Initial Treatment and Tumor Progression", p = 0.47, levels = c("TURB+BCG", "TURB+MC", "TURB alone"))
```

### Cyclin A1 and Tumor Progression (Figure 2H)
```{r Figure2H, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.progression ~ ca1.positive.10)
# Survival curves
survival.plot(x = ca1.progression, title = "Cyclin A1 and Tumor Progression", p = 0.004, levels = c("Negative", "Positive"))
```

### Tumor Progression (All Figures Combined)
```{r Figure2EH, fig.height = 11, fig.width = 13}
par(mfrow = c(2,2))
# Figure 2E
survival.plot(x = pt1.progression, title = "pT at Biopsy and Tumor Progression", p = 1.11e-06, levels = c("pT1a", "pT1b"))
# Figure 2F
survival.plot(x = cis.progression, title = "CIS at Biopsy and Tumor Progression", p = 0.03, levels = c("Absent", "Present"), position = "bottomright")
# Figure 2G
survival.plot(x = treatment.progression, title = "Initial Treatment and Tumor Progression", p = 0.47, levels = c("TURB+BCG", "TURB+MC", "TURB alone"))
# Figure 2H
survival.plot(x = ca1.progression, title = "Cyclin A1 and Tumor Progression", p = 0.004, levels = c("Negative", "Positive"))

```

## Estimating Hazard Ratios for Tumor Recurrence
cox.pt1.recurrence <- coxph(surv.recurrence ~ pt.biopsy, data = ca1) # pT1 and recurrence
cox.cis.recurrence <- coxph(surv.recurrence ~ cis.biopsy, data = ca1) # CIS and recurrence
cox.treatment.recurrence <- coxph(surv.recurrence ~ treatment, data = ca1) # Initial treatment and recurrence
cox.ca1.recurrence <- coxph(surv.recurrence ~ ca1.positive.10, data = ca1) # Cyclin A1 and recurrence
cox.ca1pt1.recurrence <- coxph(surv.recurrence ~ ca1.positive.10 + pt.biopsy, data = ca1) # Cyclin A1 & pT1 at Biopsy and recurrence


## Estimating Hazard Ratios for Tumor Progression
cox.pt1.progression <- coxph(surv.progression ~ pt.biopsy, data = ca1) # pT1 and progression
cox.cis.progression <- coxph(surv.progression ~ cis.biopsy, data = ca1) # CIS and progression
cox.treatment.progression <- coxph(surv.progression ~ treatment, data = ca1) # Initial treatment and progression
cox.ca1.progression <- coxph(surv.progression ~ ca1.positive.10, data = ca1) # Cyclin A1 and progression
cox.ca1pt1.progression <- coxph(surv.progression ~ ca1.positive.10 + pt.biopsy, data = ca1) # Cyclin A1 & pT1 at Biopsy and progression

# Table 2
# Tumor Recurrence
summary(cox.pt1.recurrence) # pT1 and recurrence
summary(cox.cis.recurrence) # CIS and recurrence
summary(cox.treatment.recurrence) # Treatment and recurrence
summary(cox.ca1.recurrence) # Cyclin A1 and recurrence
summary(cox.ca1pt1.recurrence) # Cyclin A1 & pT1 and recurrence
# Tumor Progression
summary(cox.pt1.progression) # pT1 and progression
summary(cox.cis.progression) # CIS and progression
summary(cox.treatment.progression) # Treatment and progression
summary(cox.ca1.progression) # Cyclin A1 and progression
summary(cox.ca1pt1.progression) # Cyclin A1 & pT1 and progression
# Evaluating the interaction between pT1 and Cyclin A1
cox.ca1bypt1.progression <- coxph(surv.progression ~ ca1.positive.10*pt.biopsy, data = ca1) 
summary(cox.ca1bypt1.progression)
# Comparison of Model 1 (only pT1) vs. Model 2 (pT1 + Cyclin A1)
# vs. Model 3 (only Cyclin A1) for tumor progression
anova(cox.pt1.progression, cox.ca1pt1.progression) # Model 1 vs. Model 2
anova(cox.ca1.progression, cox.ca1pt1.progression) # Model 3 vs. Model 2