---
output:
  html_document: default
---
## Cyclin A1 Expression in pT1 Urothelial Carcinoma of Bladder
* Opening & attaching dataset and loading functions & libraries
```{r open}
ca1 <- read.csv("../ca1.csv")
attach(ca1)
library(psych) # For descriptive statistics
library(survival) # For Survival Analyses
source("../standardAnalysis.R")
source("../SurvivalPlot.R")
```
### Describing the Data
```{r describe}
# pT1 at Biopsy
table.prop(pt.biopsy)
# CIS at Biopsy
table.prop(cis.biopsy) 
# Initial Treatment
table.prop(treatment)
# Recurrence ratio
table.prop(recurrence)
# Distribution of TNM stages
table.prop(dx.final) 
# Describing the number of recurrence episodes
descriptive(recurrence.number) 
# Recurrence episodes by groups
table.prop(recurrence.groups) 
# Describing the % of cyclin A1 expression
descriptive(ca1.extension)
# % Cyclin A1 Positivity
table.prop(ca1.positive.10)
```

### Table 1: Cyclin A1 Immunohistochemical Expression in 149 Invasive High-Grade Urothelial Carcinomas by Pathologic Features and Outcome
```{r}
# Cyclin A1 by pT at Biopsy
compare.fisher(pt.biopsy, ca1.positive.10)
# Cyclin A1 by CIS at Biopsy
compare.fisher(cis.biopsy, ca1.positive.10)
# Cyclin A1 by Recurrence
compare.fisher(recurrence, ca1.positive.10)
# Cyclin A1 by Progression
compare.fisher(progression, ca1.positive.10) 
# Adjusting P value using Hommel's correction
table1.raw <- c(0.01864, 0.7189, 1, 0.00325)
table1.hommel <- p.adjust(table1.raw, method = c("hommel"))
table1.adjusted <- cbind(table1.raw, table1.hommel)
matrix(data = table1.adjusted, nrow = 4, ncol = 2,
       dimnames = list(c("pT at Biopsy", "CIS at Biopsy", "Recurrence", "Progression"),
                       c("Raw P values", "Adjusted P values")))
```

### Survival Plots for Tumor Recurrence
* Estimating Survival Functions for Tumor Recurrence
```{r SurvivalRecurrence}
# Defining survival object
surv.recurrence <- with(ca1, Surv(time.event, recurrence))
# pT1 and recurrence
pt1.recurrence <- survfit(surv.recurrence ~ pt.biopsy, data = ca1)
# CIS and recurrence
cis.recurrence <- survfit(surv.recurrence ~ cis.biopsy, data = ca1)
# Initial treatment and recurrence
treatment.recurrence <- survfit(surv.recurrence ~ treatment, data = ca1)
# Cyclin A1 and recurrence
ca1.recurrence <- survfit(surv.recurrence ~ ca1.positive.10, data = ca1)
```

* pT Biopsy and Tumor Recurrence (Figure 2A)
```{r Figure2A, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.recurrence ~ pt.biopsy)
# Survival curves
survival.plot(x = pt1.recurrence, title = "pT at Biopsy and Tumor Recurrence",
p = 0.02, levels = c("pT1a", "pT1b"))
```

* CIS and Tumor Recurrence (Figure 2B)
```{r Figure2B, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.recurrence ~ cis.biopsy)
# Survival curves
survival.plot(x = cis.recurrence, title = "CIS at Biopsy and Tumor Recurrence",
p = 0.32, levels = c("Absent", "Present"))
```

* Initial Treatment and Tumor Recurrence (Figure 2C)
```{r Figure2C, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.recurrence ~ treatment)
# Survival curves
survival.plot(x = treatment.recurrence, title = "Initial Treatment at Biopsy and Tumor Recurrence", p = 0.47, levels = c("TURB+BCG", "TURB+MC", "TURB alone"))
```

* Cyclin A1 and Tumor Recurrence (Figure 2D)
```{r Figure2D, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.recurrence ~ ca1.positive.10)
# Survival curves
survival.plot(x = ca1.recurrence, title = "Cyclin A1 and Tumor Recurrence", p = 0.68, position = "topright", levels = c("Negative", "Positive"))
```

* Tumor Recurrence (All Figures Combined)
```{r Figure2AD, fig.height = 11, fig.width = 13}
par(mfrow = c(2,2))
# Figure 2A
survival.plot(x = pt1.recurrence, title = "pT at Biopsy and Tumor Recurrence",
p = 0.02, levels = c("pT1a", "pT1b"))
# Figure 2B
survival.plot(x = cis.recurrence, title = "CIS at Biopsy and Tumor Recurrence",
p = 0.32, levels = c("Absent", "Present"))
# Figure 2C
survival.plot(x = treatment.recurrence, title = "Initial Treatment at Biopsy and Tumor Recurrence", p = 0.47, levels = c("TURB+BCG", "TURB+MC", "TURB alone"))
# Figure 2D
survival.plot(x = ca1.recurrence, title = "Cyclin A1 and Tumor Recurrence", p = 0.68, position = "topright", levels = c("Negative", "Positive"))
```

### Survival Plots for Tumor Progression
* Estimating Survival Functions for Tumor Progression
```{r SurvivalProgression}
# For tumor progression
surv.progression <- with(ca1, Surv(time.event, progression))
# pT1 and progression
pt1.progression <- survfit(surv.progression ~ pt.biopsy, data = ca1)
# CIS and progression
cis.progression <- survfit(surv.progression ~ cis.biopsy, data = ca1)
# Initial treatment and progression
treatment.progression <- survfit(surv.progression ~ treatment, data = ca1)
# Cyclin A1 and progression
ca1.progression <- survfit(surv.progression ~ ca1.positive.10, data = ca1)
```

* pT at Biopsy and Tumor Progression (Figure 2E)
```{r Figure2E, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.progression ~ pt.biopsy)
# Survival curves
survival.plot(x = pt1.progression, title = "pT at Biopsy and Tumor Progression", p = 1.11e-06, levels = c("pT1a", "pT1b"))
```

* CIS at Biopsy and Tumor Progression (Figure 2F)
```{r Figure2F, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.progression ~ cis.biopsy)
# Survival curves
survival.plot(x = cis.progression, title = "CIS at Biopsy and Tumor Progression", p = 0.03, levels = c("Absent", "Present"), position = "bottomright")
```

* Initial Treatment and Tumor Progression (Figure 2G)
```{r Figure2G, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.progression ~ treatment)
# Survival curves
survival.plot(x = treatment.progression, title = "Initial Treatment and Tumor Progression", p = 0.47, levels = c("TURB+BCG", "TURB+MC", "TURB alone"))
```

* Cyclin A1 and Tumor Progression (Figure 2H)
```{r Figure2H, fig.height = 7, fig.width = 9}
# Log-rank test
survdiff(surv.progression ~ ca1.positive.10)
# Survival curves
survival.plot(x = ca1.progression, title = "Cyclin A1 and Tumor Progression", p = 0.004, levels = c("Negative", "Positive"))
```

* Tumor Progression (All Figures Combined)
```{r Figure2EH, fig.height = 11, fig.width = 13}
par(mfrow = c(2,2))
# Figure 2E
survival.plot(x = pt1.progression, title = "pT at Biopsy and Tumor Progression", p = 1.11e-06, levels = c("pT1a", "pT1b"))
# Figure 2F
survival.plot(x = cis.progression, title = "CIS at Biopsy and Tumor Progression", p = 0.03, levels = c("Absent", "Present"), position = "bottomright")
# Figure 2G
survival.plot(x = treatment.progression, title = "Initial Treatment and Tumor Progression", p = 0.47, levels = c("TURB+BCG", "TURB+MC", "TURB alone"))
# Figure 2H
survival.plot(x = ca1.progression, title = "Cyclin A1 and Tumor Progression", p = 0.004, levels = c("Negative", "Positive"))
```

### pT Substaging and Cyclin A1 Expression in Predicting Tumor Recurrence and Tumor Progression
First we create a new variable containing the 4 possible combinations of pT substage and cyclin A1 expression (i.e., pT1a/CycA1+, pT1a/CycA1-, pT1b/CycA1+, pT1b/CycA1-) and we plot the proportions of each level:
```{r, echo = TRUE}
ca1$pt.ca1[ca1$pt.biopsy == "pT1a" & ca1$ca1.positive.10 == "Positive"] <- "pT1a/CycA1+"
ca1$pt.ca1[ca1$pt.biopsy == "pT1a" & ca1$ca1.positive.10 == "Negative"] <- "pT1a/CycA1-"
ca1$pt.ca1[ca1$pt.biopsy == "pT1b" & ca1$ca1.positive.10 == "Positive"] <- "pT1b/CycA1+"
ca1$pt.ca1[ca1$pt.biopsy == "pT1b" & ca1$ca1.positive.10 == "Negative"] <- "pT1b/CycA1-"
table.prop(ca1$pt.ca1)
```
We then plot the survival curves for tumor recurrence and tumor progression considering these 4 levels.

* Combined pT1 substaging and cyclin A1 expresion in predicting tumor recurrence
```{r Figure3A, fig.height = 7, fig.width = 9}
# Defining survival function
ptca1.recurrence <- survfit(surv.recurrence ~ pt.ca1, data = ca1)
# Log-rank test
with(ca1, survdiff(surv.recurrence ~ pt.ca1))
# Figure 3A
survival.plot(x = ptca1.recurrence, title = "pT1 Substage and Cyclin A1 for Predicting \nTumor Recurrence", p = 0.014, levels = c("pT1a/CycA1-", "pT1a/CycA1+", "pT1b/CycA1-", "pT1b/CycA1+"))
```

* Combined pT1 substaging and cyclin A1 expresion in predicting tumor progression
```{r Figure3B, fig.height = 7, fig.width = 9}
# Defining survival function
ptca1.progression <- survfit(surv.progression ~ pt.ca1, data = ca1)
# Log-rank test
with(ca1, survdiff(surv.progression ~ pt.ca1))
# Figure 3B
survival.plot(x = ptca1.progression, title = "pT1 Substage and Cyclin A1 for Predicting \nTumor Progression", p = 7.7e-07, levels = c("pT1a/CycA1-", "pT1a/CycA1+", "pT1b/CycA1-", "pT1b/CycA1+"))
```

### Table 2: Hazard Ratios for Tumor Recurrence and Tumor Progression According to Pathologic Features at Biopsy and Cyclin A1 Immunohistochemical Expression in 149 Invasive High-Grade Urothelial Carcinomas
* Hazard ratios for tumor recurrence and tumor progression
```{r HR}
# pT1 and recurrence
hazard(surv.recurrence ~ pt.biopsy, data = ca1)
# CIS and recurrence
hazard(surv.recurrence ~ cis.biopsy, data = ca1)
# Initial treatment and recurrence
hazard(surv.recurrence ~ treatment, data = ca1)
# Cyclin A1 and recurrence
hazard(surv.recurrence ~ ca1.positive.10, data = ca1)
# Cyclin A1 & pT1 at Biopsy and recurrence
hazard(surv.recurrence ~ ca1.positive.10 + pt.biopsy, data = ca1)
# pT1 and progression
hazard(surv.progression ~ pt.biopsy, data = ca1)
# CIS and progression
hazard(surv.progression ~ cis.biopsy, data = ca1)
# Initial treatment and progression
hazard(surv.progression ~ treatment, data = ca1)
# Cyclin A1 and progression
hazard(surv.progression ~ ca1.positive.10, data = ca1)
# Cyclin A1 & pT1 at Biopsy and progression
hazard(surv.progression ~ ca1.positive.10 + pt.biopsy, data = ca1)
# Interaction between pT1 and Cyclin A1
hazard(surv.progression ~ ca1.positive.10*pt.biopsy, data = ca1) 
```

### Comparison of Model 1 (only pT1) vs. Model 2 (pT1 + Cyclin A1) vs. Model 3 (only Cyclin A1) for tumor progression
```{r}
model1 <- coxph(surv.progression ~ pt.biopsy, data = ca1)
model2 <- coxph(surv.progression ~ pt.biopsy + ca1.positive.10, data = ca1)
model3 <- coxph(surv.progression ~ ca1.positive.10, data = ca1)
# Model 1 vs. Model 2
anova(model1, model2)
# Model 3 vs. Model 2
anova(model3, model2) 
```